---
title: "NCAA Men's March Madness"
description: |
  Exploring historical Final Four success and public championship picks
author: Matthew Diekmann
date: September 28, 2026
format:
  html:
    code-fold: true
execute:
  warning: false
  message: false
---

## About the Data

For this analysis, I used the NCAA Men's March Madness dataset featured in
[TidyTuesday](https://github.com/rfordatascience/tidytuesday/tree/main/data/2024/2024-03-26).

The data come from Nishaan Amin's March Madness dataset and contain historical
team tournament results along with public predictions for tournament teams.

The data cover men's NCAA tournaments from 2008 through 2024, excluding 2020
because the tournament was canceled.

## Load the Data

```{r}
library(tidyverse)

team_results <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-03-26/team-results.csv"
)

public_picks <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-03-26/public-picks.csv"
)
```

## Which Teams Have Made the Most Final Fours?

The `F4` variable gives the number of times each program reached the Final Four.
I looked at the teams with the most Final Four appearances in the dataset.

```{r}
top_final_four <- team_results |>
  filter(F4 > 0) |>
  arrange(desc(F4)) |>
  slice_head(n = 15)

ggplot(
  top_final_four,
  aes(
    x = reorder(TEAM, F4),
    y = F4
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Teams With the Most Final Four Appearances",
    subtitle = "NCAA men's tournament results in the TidyTuesday dataset",
    x = "Team",
    y = "Final Four appearances"
  ) +
  theme_minimal()
```

This visualization shows which programs have reached the Final Four most often
over the period represented in the dataset.

## Which Teams Did the Public Most Expect to Win in 2024?

The `public_picks` dataset contains the percentage of public brackets that
picked each team to advance through each round of the 2024 tournament. The
`FINALS` variable gives the percentage of people who picked each team to win
the championship.

I looked at the 10 teams with the highest championship pick percentages.

```{r}
top_public_picks <- public_picks |>
  mutate(
    finals_pct = readr::parse_number(FINALS)
  ) |>
  filter(!is.na(finals_pct)) |>
  arrange(desc(finals_pct)) |>
  slice_head(n = 10)

ggplot(
  top_public_picks,
  aes(
    x = reorder(TEAM, finals_pct),
    y = finals_pct
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Most Popular Public Picks to Win the 2024 Championship",
    subtitle = "Top 10 teams by percentage of public brackets",
    x = "Team",
    y = "Percent of public brackets picking team to win"
  ) +
  theme_minimal()
```

Connecticut was the most popular public choice to win the championship in
2024, while several other highly rated teams also received substantial support.
This visualization shows how concentrated public expectations were among the
top contenders.

As a side note, UCONN went on to have one of the most dominant title runs in March Madness history, winning by a record average margin of 23 points per game.

## Sources

**TidyTuesday:**  
[TidyTuesday — NCAA Men's March Madness](https://github.com/rfordatascience/tidytuesday/tree/main/data/2024/2024-03-26)

**Original data:**  
Amin, Nishaan. *March Madness Data*. Kaggle.
[Original March Madness dataset](https://www.kaggle.com/datasets/nishaanamin/march-madness-data)