---
title: "Billboard Hot 100 Number Ones"
description: |
  Exploring song length and genre among Billboard Hot 100 number-one hits
author: Matthew Diekmann
date: September 28, 2026
format:
  html:
    code-fold: true
execute:
  warning: false
  message: false
---

## About the Data

For this analysis, I used the Billboard Hot 100 Number Ones dataset featured in
[TidyTuesday](https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-08-26).

The dataset contains information about songs that reached number one on the
Billboard Hot 100 between 1958 and January 2025. It includes information about
the songs' artists, genres, length, musical characteristics, and chart
performance.

The original dataset was compiled by Chris Dalla Riva as part of his
[Billboard Hot 100 Number Ones Database](https://docs.google.com/spreadsheets/d/1j1AUgtMnjpFTz54UdXgCKZ1i4bNxFjf01ImJ-BqBEt0/edit).

## Load the Data

```{r}
library(tidyverse)

billboard <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-08-26/billboard.csv"
)

topics <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-08-26/topics.csv"
)
```

## Have Number-One Songs Gotten Shorter?

One question I was interested in was whether the length of number-one songs has
changed over time. The dataset includes `length_sec`, which measures the length
of each song in seconds. I converted this to minutes to make the visualization
easier to interpret.

```{r}
billboard_length <- billboard |>
  mutate(
    year = as.integer(format(date, "%Y")),
    length_min = length_sec / 60
  ) |>
  filter(!is.na(length_min))

ggplot(
  billboard_length,
  aes(x = year, y = length_min)
) +
  geom_point(alpha = 0.35) +
  geom_smooth(se = FALSE, method = "loess") +
  labs(
    title = "Length of Billboard Hot 100 Number-One Songs Over Time",
    subtitle = "Each point represents a song that reached number one",
    x = "Year",
    y = "Song length (minutes)"
  ) +
  theme_minimal()
```

The trend line makes it easier to see how the typical length of a number-one
song has changed over time. Song lengths increased during parts of the late
20th century, but more recent number-one hits appear to have become shorter.

## Which Genres Have the Most Number-One Hits?

I also wanted to see which genres appear most frequently among number-one
songs. The variable `cdr_genre` gives the genre assigned to each song in the
dataset.

```{r}
top_genres <- billboard |>
  filter(!is.na(cdr_genre)) |>
  count(cdr_genre, sort = TRUE) |>
  slice_head(n = 10)

ggplot(
  top_genres,
  aes(
    x = reorder(cdr_genre, n),
    y = n
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Genres With the Most Billboard Hot 100 Number-One Songs",
    x = "Genre",
    y = "Number of number-one songs"
  ) +
  theme_minimal()
```

This visualization shows which genres have accounted for the largest number of
Billboard Hot 100 number-one songs in the dataset.

## Sources

**TidyTuesday:**  
[TidyTuesday — Billboard Hot 100 Number Ones](https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-08-26)

**Original data:**  
Dalla Riva, Chris. *Billboard Hot 100 Number Ones Database*.  
[Billboard Hot 100 Number Ones Database](https://docs.google.com/spreadsheets/d/1j1AUgtMnjpFTz54UdXgCKZ1i4bNxFjf01ImJ-BqBEt0/edit)